Speed and Precision in Range Analysis

### 2012-07-01 doc/Manuscript/SBLP/paper.tex
My revision of the paper.

% Define the lattice, the constraints and the valuation I.
\noindent
\textbf{The Interval Lattice.}
Following Gawlitza {\em et al.}'s notation, we shall be performing arithmetic
operations over the lattice
$\cal{Z} = \mathbb{Z} \cup \{-\infty, +\infty\}$, where the ordering is
naturally given by $-\infty < \ldots -1 < 0 < 1 \ldots +\infty$.
We let meet and join be the min and max operators respectively.
For any $x > -\infty$ we define:

@@

\noindent
\textbf{A Holistic View of our Range Analysis Algorithm.}
Figure~\ref{fig:algorithm} gives a global view of our range analysis
algorithm.
We perform range analysis in a number of steps, some of which are optional.
The optional parts improve the precision of the range analysis, at the expense of
a longer running time.
In Section~\ref{sec:design} we discuss in more details these tradeoffs.

@@

As we see from a cursory glance at Figure~\ref{fig:algorithm}, our range
analysis algorithm has many optional modules.
These modules give the user the chance to choose between more precise results,
or a faster analysis.
Given the number of options, the design space of a range analysis algorithm
is vast.
In this section we try to cover some of the most important tradeoffs.
All the numbers that we show have been obtained as the average of 15 runs in
an Intel Duo CPU with a 3.00GHz clock, and 2.04GHz main memory.
Figure~\ref{fig:space} plots, for the integer programs in the SPEC CPU 2006
benchmark suite, precision versus speed for different configurations of
our implementation.
Our initial goal when developing this analysis was to support a bitwidth-aware
register allocator.
Thus, we measure precision by the average number of bits that our
analysis allows us to save per program variable.
It is very important to notice that we do not consider constants in our statistics of precision.
In other words, we only measure bitwidth reduction in variables that a constant
propagation step could not remove.

@@

According to Figure~\ref{fig:impactSCC}, in some cases, as in \texttt{bzip2},
strong components increase our precision by 40\%.
The gains in precision happen because, by completely resolving a component,
we are able to propagate constant intervals to the next components, instead
of propagating intervals that can grow in both directions.
An an example, in Figure~\ref{fig:ex_partition_grow_crop} we pass the range
$[0, 99]$ from variable $k$ to the component that contains variable $j$.
Had we run the analysis in the entire constraint graph, by the time we
applied the growth analysis on $j$ we would still find $k$ bound to
$0, +\infty$.

@@

Figure~\ref{fig:impactESSA} compares these two program representations in
terms of runtime.
As we see in Figure~\ref{fig:impactESSA}(Left), the e-SSA form slows down our
analysis.
In some cases, as in \texttt{xalancbmk}, this slowdown increases execution
time by 71\%.
Runtime increases for two reasons.
Firstly, the e-SSA form programs are larger than the SSA form programs, as we
show in Figure~\ref{fig:impactESSA}(Right).
However, this growth is small: we did not verify any growth larger than 9\% in
any integer program of SPEC CPU 2006.
Second, the e-SSA form program has futures; hence requiring the future resolution
phase of our algorithm, which is not necessary in SSA form programs.
Nevertheless, if the e-SSA form slow downs the analysis runtime, its gains in
precision are remarkable, as seen in Figure~\ref{fig:precESSA}.
These gains happen because the e-SSA format lets the analysis to use the
results of comparisons to narrow the ranges of variables.

@@

Figure~\ref{fig:wholeImpact} compares the precision of the intra and
inter-procedural analyses for the five largest programs in three different
categories of benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
Bitwise~\cite{Stephenson00}.
Our results for the SPEC programs were disappointing: on average, for
the five largest programs, the intra-procedural version of our analysis saves
5.23\% of bits per variable.
The inter-procedural version increases this number to 8.89\%.
A manual inspection of the SPEC programs reveals that this result is expected:
these programs use may external library functions, which we cannot analyze, and
their source codes do no provide enough explicit constants to power our analysis
up.
However, with numerical benchmarks we fare much better.
On average, our inter-procedural algorithm reduces the bitwidth of the
Stanford benchmarks by 36.24\%.
For Bitwise we obtain a bitwidth reduction of 12.27\%.
However, this average is lowered by two outliers: \texttt{edge\_detect} and
\texttt{sha}, which cannot be reduced.
The bitwise benchmarks were implemented by Stephenson
{\em et al.}~\cite{Stephenson00} to validate their bitwidth analysis.
Our results are on par with those found by the original authors.
The bitwise programs contain only the \texttt{main} function; thus, different
versions of our algorithm find the same results when applied onto these
programs.

@@

Another way to increase the precision of range analysis is via a
context-sensitive implementation.
Context-sensitiveness allows us to distinguish different calling sites of
the same function.
Figure~\ref{fig:context} shows why the ability to make this distinction is
important for precision.
In Figure~\ref{fig:context}(a) we have two different calls of function
\texttt{foo}.
An usual way to perform a data-flow analysis inter-procedurally is to
create assignments between formal and actual parameters, as we show in
Figure~\ref{fig:context}(b).
If a function is called more than once, then its formal parameters will receive
information from many actual parameters.
We use the SSA's $\phi$-functions to bind this information together into a single
flow.
However, in this case the multiple assignment of values to parameters
makes the ranges of these parameters very large, whereas in reality they
are not.
As an example, in Figure~\ref{fig:context}(b), variable \texttt{k} ends up
associated with the range $[0, 10^5]$, but in reality this variable contains
an interval that is only 100 units long.
A way to circumvent this source of imprecision is via function inlining,
as we show in Figure~\ref{fig:context}(c).
The results that we can derive for the transformed program are more
precise, as each input parameter is assigned a single value.

@@

Figure~\ref{fig:wholeImpact} shows how function inlining modifies the
precision of our results.
It is difficult to find an adequate way to compare the precision of
our analysis with, and without inlining.
This difficulty stems from the fact that this transformation tends to change
the target program too much.
In absolute numbers, we always reduce the bitwidth of more variables after
function inlining.
However, proportionally function inlining leads to a smaller percentage of
bitwidth reduction for many benchmarks.
In the Stanford Collection, for instance, where most of the functions are
called in only one location, inlining leads to worst precision results.
On the other hand, for the SPEC programs, inlining, even in terms of
percentage of reduction, tends to increase our measure of precision.

>>>>

% Define the lattice, the constraints and the valuation I.
\noindent
\textbf{The Interval Lattice.}
Following Gawlitza {\em et al.}'s notation, we shall be performing arithmetic
operations over the lattice
$\cal{Z} = \mathbb{Z} \cup \{-\infty, +\infty\}$, where the ordering is
naturally given by $-\infty < \ldots -1 < 0 < 1 \ldots < +\infty$.
We let meet and join be the min and max operators respectively.
For any $x > -\infty$ we define:

@@

\noindent
\textbf{A Holistic View of our Range Analysis Algorithm.}
Figure~\ref{fig:algorithm} gives a global view of our range analysis
algorithm.
We perform range analysis in a number of steps, some of which are optional.
The optional parts improve the precision of the range analysis, at the expense of
a longer running time.
In Section~\ref{sec:design} we discuss in more detail these tradeoffs.

@@

As we see from a cursory glance at Figure~\ref{fig:algorithm}, our range
analysis algorithm has many optional modules.
These modules give the user the chance to choose between more precise results,
or a faster analysis.
Given the number of options, the design space of a range analysis algorithm
is vast.
In this section we try to cover some of the most important tradeoffs.
All the numbers that we show have been obtained as the average of 15 runs in
an Intel Core 2 Quad processor with 2.4 GHz, and 3.5 GB of main memory.
Figure~\ref{fig:space} plots, for the integer programs in the SPEC CPU 2006
benchmark suite, precision versus speed for different configurations of
our implementation.
Our initial goal when developing this analysis was to support a bitwidth-aware
register allocator.
Thus, we measure precision by the average number of bits that our
analysis allows us to save per program variable.
It is very important to notice that we do not consider constants in our statistics of precision.
In other words, we only measure bitwidth reduction in variables that a constant
propagation step could not remove.

@@

According to Figure~\ref{fig:impactSCC}, in some cases, as in \texttt{bzip2},
strong components increase our precision by 40\%.
The gains in precision happen because, by completely resolving a component,
we are able to propagate constant intervals to the next components, instead
of propagating intervals that can grow in both directions.
An an example, in Figure~\ref{fig:ex_partition_grow_crop} we pass the range
$[0, 99]$ from variable $k$ to the component that contains variable $j$.
Had we run the analysis in the entire constraint graph, by the time we
applied the growth analysis on $j$ we would still find $k$ bound to
$[0, +\infty]$.

@@

Figure~\ref{fig:impactESSA} compares these two program representations in
terms of runtime.
As we see in Figure~\ref{fig:impactESSA}(Left), the e-SSA form slows down our
analysis.
In some cases, as in \texttt{xalancbmk}, this slowdown increases execution
time by 71\%.
Runtime increases for two reasons.
Firstly, the e-SSA form programs are larger than the SSA form programs, as we
show in Figure~\ref{fig:impactESSA}(Right).
However, this growth is small: we did not verify any growth larger than 9\% in
any integer program of SPEC CPU 2006.
Second, the e-SSA form program has futures; hence requiring the future resolution
phase of our algorithm, which is not necessary in SSA form programs.
Nevertheless, whereas the e-SSA form slows down the analysis runtime, its gains in
precision are remarkable, as seen in Figure~\ref{fig:precESSA}.
These gains happen because the e-SSA format lets the analysis to use the
results of comparisons to narrow the ranges of variables.

@@

Figure~\ref{fig:wholeImpact} compares the precision of the intra and
inter-procedural analyses for the five largest programs in three different
categories of benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
Bitwise~\cite{Stephenson00}.
Our results for the SPEC programs were disappointing: on average, for
the five largest programs, the intra-procedural version of our analysis saves
5.23\% of bits per variable.
The inter-procedural version increases this number to 8.89\%.
A manual inspection of the SPEC programs reveals that this result is expected:
these programs use many external library functions, which we cannot analyze, and
their source codes do not provide enough explicit constants to power our analysis
up.
However, with numerical benchmarks we fare much better.
On average, our inter-procedural algorithm reduces the bitwidth of the
Stanford benchmarks by 36.24\%.
For Bitwise we obtain a bitwidth reduction of 12.27\%.
However, this average is lowered by two outliers: \texttt{edge\_detect} and
\texttt{sha}, which cannot be reduced.
The bitwise benchmarks were implemented by Stephenson
{\em et al.}~\cite{Stephenson00} to validate their bitwidth analysis.
Our results are on par with those found by the original authors.
The bitwise programs contain only the \texttt{main} function; thus, different
versions of our algorithm find the same results when applied onto these
programs.

@@

Another way to increase the precision of range analysis is via a
context-sensitive implementation.
Context-sensitiveness allows us to distinguish different calling sites of
the same function.
Figure~\ref{fig:context} shows why the ability to make this distinction is
important for precision.
In Figure~\ref{fig:context}(a) we have two different calls of function
\texttt{foo}.
An usual way to perform a data-flow analysis inter-procedurally is to
create assignments between formal and actual parameters, as we show in
Figure~\ref{fig:context}(b).
If a function is called more than once, then its formal parameters will receive
information from many actual parameters.
We use the SSA's $\phi$-functions to bind this information together into a single
flow.
However, in this case the multiple assignment of values to parameters
makes the ranges of these parameters very large, whereas in reality they
are not.
As an example, in Figure~\ref{fig:context}(b), variable $k$ ends up
associated with the range $[0, 10^5]$, but in reality this variable contains
an interval that is only 100 units long.
A way to circumvent this source of imprecision is via function inlining,
as we show in Figure~\ref{fig:context}(c).
The results that we can derive for the transformed program are more
precise, as each input parameter is assigned a single value.

@@

Figure~\ref{fig:wholeImpact} shows how function inlining modifies the
precision of our results.
It is difficult to find an adequate way to compare the precision of
our analysis with, and without inlining.
This difficulty stems from the fact that this transformation tends to change
the target program too much.
In absolute numbers, we always reduce the bitwidth of more variables after
function inlining.
However, proportionally function inlining leads to a smaller percentage of
bitwidth reduction for many benchmarks.
In the Stanford Collection, for instance, where most of the functions are
called in only one location, inlining leads to worse precision results.
On the other hand, for the SPEC programs, inlining, even in terms of
percentage of reduction, tends to increase our measure of precision.
